---
title: "Estimate caribou parturition using the TuktuTools Package"
author: "Ophélie Couriot"
date: "12/14/2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette_parturition}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, echo = FALSE,
                      rmarkdown.html_vignette.check_title = FALSE)
```


# Introduction

This document provides steps to estimate calving (i.e., parturition) status, timing and location, for female barren-ground caribou (*Rangifer tarandus spp.*), by analyzing their movement patterns using functions from the `TuktuTools` package. 

We adapted the individual-based method developed by DeMars et al (2013) to infer calving status. The DeMars et al. (2013) method assumes for calving, a female decreases her movements and that the distance traveled after calving remains low for several days or weeks, as long as the calf is alive, since neonates are not able to move as quickly adults. Thus, calving events are identifiable by a sudden and marked change – or break point – in a female's mean step length (i.e. the distance between two successive relocations). Conversely, if the female loses her calf during the neonatal period, a second break point would occur in the female movement pattern, since the female would recover her pre-calving movement rate abruptly, due to calf loss. The method also assumes that contrary to calving females, non-calving females have a constant movement throughout the entire calving period. This approach requires an homogeneous time series (i.e. homogeneous time interval between successive relocations). To account for variability in sample schedules and missing data, we adapted this method by performing analyses on movement rate (i.e. the distance traveled between two successive relocations) instead of step length. More details on this method can be found in Couriot et al. (In press), and users are encouraged to read the paper for a deeper understanding of the methodology. 


# TuktuTools

The `TuktuTools` package contains a selection of tools to:

* Clean and process movement data
* Visualize movement data
* Analyze spatial patterns
    + Estimate individual movement rate
    + Calculate euclidean distance between pairs of individuals
    + Estimate ranges of an individual or all individuals during a given period using two different methods (LoCoH or KUD)
    + Discriminate between parturient and non parturient individuals and identify parturition timing and location

In this document, we focus on the `estimateCalving` function, which estimates calving status, timing and location, using an individual based method. We provide the required steps to estimate calving, from data processing and cleaning to identifying calving timing and location. In order to provide concrete examples, the package includes movement data from four females of barren-ground caribou, provided by the Government of the Northwest Territories under the agreement *8515-05 DR #607*.

# Estimate calving

```{r LoadLibraries, echo = FALSE, eval = TRUE, results = 'hide'}
require(TuktuTools)
require(kableExtra)
require(ggplot2)
require(ggthemes)
require(ggpubr)
require(ggrepel)
require(mapview)
```

## Clean data

The GPS devices are configured to relocate daily, every 8-hours, or hourly for some individuals (e.g. 0:00 am, 8:00 am or 4:00 pm). However, it appears that sometimes the GPS devices relocate at 1 or 2 minute intervals not fitting with the hourly schedule. In addition, it is biologically impossible for an animal to move several kilometers in one minute. Thus, the relocations with a time interval in the order of minutes are more prone to be 'outliers'.

The `removeOutliers` function in this package allows users to clean data prior to analyses. The function flags relocations that are not biologically probable, using the following rules:

* if speed between the fixes is higher than 50 km per hour
* if time interval between fixes is less than 2 minutes
* if speed is higher than 20 km per hour and the time interval is less than 10 minutes

### Usage

```{r, echo = TRUE, eval = FALSE}
removeOutliers(df, steps = 10)
```


### Arguments

`df`   a data frame containing columns: ID - individual identifier, x and y coordinates (in a metric system), DateTime - date and time vector (of class POSIXct)

`steps`   if specified, the number of cleaning steps to be performed (default is 10)


### Value

The function adds a column "outlier". If TRUE, it means that the location has been identified as an outlier.

### Application

```{r removeOutliers, echo = TRUE, eval = TRUE}
# load data
data(caribou)

# remove potential outliers and keep only females
caribou.cleaned <- removeOutliers(caribou) %>% subset(outlier == "FALSE" & sex == "f")
```


## Focus on the calving period

After removing potential outliers, we censored and trimmed our data to focus on the calving period, to avoid detecting behaviors other than calving. The calving period has been described to occur between May 19 and July 07 for barren-ground caribou (Cameron et al. 2018). 
The convenient function `prepData` allows trimming the timeseries to a given period, guaranteeing a minimum number of location fixes a day, and removing individuals with gaps in the data greater than a threshold.

### Usage
```{r, echo = TRUE, eval = FALSE}
prepData(df, start, end, nfixes, dayloss)
```

### Arguments

`df`   a data frame containing columns: ID_Year - individual identifier per year, x and y coordinates (in a metric system), DateTime - date and time vector (of class POSIXct)

`start` period of interest start date as a character in the form "mm-dd" (example for the 19th of May: "05-19")

`end`   period of interest end date as a character in the form "mm-dd" (example for the 7th of July: "07-07")

`nfixes`   minimum number of fixes per day (taking the average number of fixes per day for each individual across the period of interest)

`dayloss`   maximum number of days with missing locations (for example, if an individual has lost signal for more than 3 consecutive days, it will be excluded from the dataset)

### Value
A filtered and processed data frame containing movement data for the desired period.

### Application

We focused our analysis on the period between May 19 to July 7, with at least 1 fix per day, and no more that 3 consecutive days of missing data.

```{r prepData, echo = TRUE, eval = TRUE}
caribou.prepped <- prepData(caribou.cleaned, start = "05-19", end = "07-07", nfixes = 1, dayloss = 3)

```

## Get movement rate 

After we processed, cleaned and filtered movement data, we can analyze movement patterns and estimate calving (i.e. parturition) status, date, location and calf death date and location, if any.

As we adapted the method developed by DeMars et al. (2013), by analyzing movement rate instead of step length. We first calculate the movement rate between subsequent relocation for each individual and each year using the `getSpeed` function.

### Usage

```{r, eval = FALSE, echo = TRUE}
getSpeed(x)
```

### Arguments
`x` a data frame containing columns: ID - individual identifier, x and y coordinates (in a metric system), Time - date and time vector (of class POSIXct)


### Value

The function adds the following columns to the dataframe:
* `speed` movement rate between subsequent locations (in meters per hour) 
* `sl` step length between subsequent locations
* `dhours` an index of the cumulative time (in hours) from the first to the nth relocation
* `dt` time lag between subsequent relocation (in hours).

### Get movement rate for all herds during the calving period

The `caribou` data contains movement data for 4 barren-ground caribou females.

```{r getSpeed, echo = TRUE, eval = TRUE}
# get movement rate for the females
caribou.mr <- getSpeed(caribou.prepped)
```

## Estimate calving status and timing

We can now estimate calving status, calving date, and calf death date, if any, using the `estimateCalving` function. This function determines calving status of a female (no calf, with a calf or lost her calf), calving date and location, and the calf death date and location. 

As stated previously, we adapted the individual based method developed by Demars et al. (2013), which has proven to reliably estimate calving status and calving date for the Western Arctic Herd (Cameron et al. 2018). However, we adapted this method to infer calving based on the female's movement rate through time.
Thus, all models assume speed follows a Gamma distribution and differ in two parameters: shape
and scale, which correspond to $\frac{\overline{speed}^{2}}{var(speed)}$ and $\frac{var(speed)}{\overline{speed}}$, respectively.

The mean speed is thus equal to $shape * scale$. For the model representing:

* females that do not calve: the mean movement rate remains constant through the entire calving period. 
* females that had a calf who survived at least 4 weeks after birth: the mean movement rate is constant before calving, then abruptly drops for calving, creating a break point. After calving, the mean movement rate increases progressively following:
$(shape_{calving} * (\frac{\overline{shape} - shape_{calving}}{k}) * time) * (scale_{calving} * (\frac{\overline{scale} - scale_{calving}}{k}) * time)$
where *k* is time, defined in days, required for the calf to achieve adult movement rates. 
* females that lose calves: there is an abrupt change in the slope of the post-calving increase, creating a second break point after which the mean speed immediately recovers its pre-calving value. 

The models therefore differ in their number of parameters to estimate: the no calf model has two – shape and scale; the calf model has five – shape and scale before calving, scale at calving, k, and calving date; and the calf death model has six – shape and scale before calving, scale at calving, k, and calving and calf death dates. We used Akaike’s Information Criterion (AIC) model selection and identified the top performing model with the lowest AIC value.


### Usage
```{r, eval = FALSE, echo = TRUE}
estimateCalving(df, int, kcons, models = c("full", "calfonly"), PlotIt = FALSE, saveplot = FALSE)
```

As we could not check for model accuracy concerning the calf death model, we decided to only compare the no calf and the calf model and then used *models = "calfonly"*.


### Arguments
`df`   a dataframe containing the movement rate between subsequent relocation of individuals, as obtained using 'getSpeed'. See ?get.speed for more information on the Data requirements

`int`   integer value indicating the minimum number of days between the beginning of the time series and the first BP (calf birth), the last BP (calf death) and the end of the time series, and between the two BPs. The main reason for this constraint is that a minimum number of data points are required in each section of the time-series to be able to estimate model parameters. We recommend 9 relocations, thereby 3 days.

`kcons`   vector of the minimum and maximum time it takes the female to recover normal movement rate (in days)

`models`  either "full" to fit all three models (i.e., no calf, calf and calf death models), or "calfonly" to fit only the no calf and calf models

`PlotIt`   if TRUE the function will draw a plot of the movement rate against date with the prediction line of the best model selected (by 'AIC'), among no-calf, calf, calf death based on the actual speed of the female in function of the date

`saveplot`   if TRUE, the plot of the best model will be saved


### Value
The function returns a list of 3 data frames:
`coeffs` containing the best model for each female, selected based on AIC, the AIC of the 3 models (i.e. no calf model, calf model and calf death model) and the negative likelihood of each model.
`Par` containing the estimated parameters of all the models (0: no calf, 1: calf, 2: calf death)
`results` containing the best model, for each female, based on AIC, the calving timing and location (if the calf model is the best model), and the mortality timing and location of the calf (if the calf death is the best model).

### Application

```{r estimateParturitions, eval = TRUE, echo = TRUE}
# Will generate a sample of two different individuals each time
part <- estimateCalving(df = caribou.mr, int=3, kcons=c(5,21), models = "calfonly", PlotIt=TRUE, saveplot=FALSE)
```


The object *`part`* is a list of 3 data frames:

```{r ShowParturitionResults}
# Coefficients
kable_styling(kable(head(part$coeffs)))

# Parameters
kable_styling(kable(head(part$par)))

# results
kable_styling(kable(head(part$results)))
```
