---
title: "Getting Weather Station with the TuktuTools R Package"
author: "Chloe Beaupré and Eliezer Gurarie"
date: "5/8/2023"
output:
  html_document:
    toc: false
    toc-float: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, out.width="100%", dpi=100, message = FALSE, warnings=FALSE)

require(mapview)
require(dplyr)
require(sf)

```

# Function description

Given a known geographical location, the `TuktuTools` package contains a function to download the corresponding raw text file of observations for the closest station from "https://www.ncei.noaa.gov/pub/data/ghcn/daily/gsn/", and converts the relevant average, maximum, minimum daily temperature into a dataframe. This function also reports the distance between the given geographical location and the corresponding station: `getWeatherAtLocation`.

## Map gsn stations

The package contains a dataset containing location information for all stations. Let's first map all the stations in the Alaska region where longitude is < -130 and latitude > 60.

```{r mapStations}
require(TuktuTools)

data("gsn_stations")
gsn_stations %>% head

gsn.sf <- st_as_sf(gsn_stations %>% subset(lon < -130 & lat > 60), 
                   coords = c("lon", "lat")) %>% st_set_crs(4326)

mapview(gsn.sf[,"station"])

```

Let's define our location of interest and plot it to see which stations are nearest.

```{r}
point <- data.frame(lon = -152.5, lat = 68.25) 
point.sf <- point %>% st_as_sf(coords = c("lon", "lat"), crs = st_crs(4326))

mapview(gsn.sf[,"station"]) + mapview(point.sf, col.regions = "red")

```

## Get weather at location

Now let's run the function. It takes a few minutes but notice as it runs that it also reports the distance between the given geographical location and the corresponding station.

```{r getWeatherAtLocation}

realWeather <- getWeatherAtLocation(point$lon, point$lat, 
                          start = 2010, end = 2022) 

# plot avg temperature by year
require(ggplot2)

realWeather %>% subset(Month == 7,) %>% arrange(Day) %>% 
  ggplot(aes(Day, Tavg, col = Station)) + 
  geom_ribbon(aes(ymin = Tmin, ymax = Tmax)) + 
  geom_point() + geom_path() + facet_wrap(.~Year)

```

