---
title: "TuktuTools: Part II - Data Summaries"
subtitle: "Workshop Materials: NACW 2023 - Anchorage"
author: "Eliezer Gurarie and Ophélie Couriot "
date: "May 8, 2023"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, out.width="100%", dpi=100, message = FALSE, warnings=FALSE)
```


The `TuktuTools` packages offers a range of functions allowing to easily summarize data and estimate movement rate and range distributions. Some examples are below.  They are based off the "prepped" caribou data from Part I, which clips the movement to a short period for the analysis of parturitions.

```{r}
require(TuktuTools)
data(caribou)
caribou_sf <- st_as_sf(caribou, 
                       coords = c("Lon","Lat"), crs = 4326)
caribou_prepped <- prepData(caribou, start = "05-19", end = "07-07") %>% 
  st_as_sf(coords = c("Lon","Lat"), crs = 4326)
```


# Get the daily mean location

`getDailyMean()` takes any large movement dataset and returns the daily mean locations for each individual for each day for each year of observations.

Let's calculate the daily mean locations for individuals during the calving period.

```{r getDailyMean}
caribou_daily_mean <- getDailyMean(caribou_prepped)
caribou_daily_mean
```
This function becomes handy when we are dealing with very large dataset, to homogenize data, by having one location per day for each individual, and do further analyses not requiring a fine fix rate.


## Get the daily movement rates

The package contains a function which is calculating the movement rate between successive locations for each individual: `getSpeed`.

Calculating the movement rate between successive daily locations. Note that this function calculate movement rate for each individual separately, but if individuals are monitored several years with a break in between (as in this case, since we considered only the calving period), it is essential to apply this function to each individual and year separately!

```{r movementRate}
daily_mr <- caribou_daily_mean %>% mutate(ID_year = paste(ID, Year)) %>% 
  getSpeed(id.col = "ID_year")
head(daily_mr)
```

We can visualize the movement rate of each individual through time. To have all the individuals on the same figure, we add a column `doy` (i.e, day of year). We also need to create a column for each ID-Year, as some individuals have been monitored several years.

```{r plotMovementRate}
daily_mr <- daily_mr %>% mutate(doy = yday(Time), ID_Year = as.factor(paste(ID, Year, sep = "_")))
ggplot(data = daily_mr, aes(x = doy, y = speed, group = ID_Year, colour = ID_Year)) + geom_line()
```

## Utilization distributions 

## Kernel Densities

In the `TuktuTools` package, we added a function that allows to easily compute kernel utilization distributions, in one line of code: `getKernelUD`. This function uses the function `kernelUD` from the `adehabitatHR` package, and thus has the same parameters found in the `kernelUD` function.

```{r getKernelUD}

individuals_kernel <- caribou_daily_mean %>% 
  st_as_sf(coords=c("x", "y"), crs = 32610) %>%
  ddply(c("ID","Year"), getKernelUD, extent = 10) %>% 
  st_as_sf(crs = 32610)

head(individuals_kernel)

# visualize
mapview::mapview(individuals_kernel, zcol = "ID")
```

## Local Convex Hulls

We also added the `getLoCoH` function, which allows to compute Local Convex Hull (LoCoH).  Read more about these here: http://tlocoh.r-forge.r-project.org/.  In short, these are known to perform better than parametric kernel methods, in the context of estimating the size of home ranges and delineating geological and ecological features in home ranges.

The LoCoH method is both a generalization of the minimum convex polygon (MCP) method and essentially a non-parametric kernel method. LoCoH applies the MCP construction to a subset of data localized in space, and the local convex polygon (i.e. local hull) is constructed using a specific number of nearest neighbors (nn) of each data point, thereby producing a set of nonparametric kernels whose union is the UD.

One advantage of using LoCoH is that this method allows to consider avoided areas (or holes) in ranges.

One complication of using this method is that you need to install the `tlocoh` package which is not on CRAN, but on another repository. 

To install T-LoCoH, type one of the following commands below at the R console. If you get stuck, see the steps for manual installation.

```{r eval = FALSE}
# Windows:
install.packages("tlocoh", dependencies=TRUE, repos=c("http://R-Forge.R-project.org", "http://cran.cnr.berkeley.edu"))
require(tlocoh)
```

```{r eval = FALSE}
# Mac / LINUX:
install.packages("tlocoh", dependencies=TRUE, repos=c("http://R-Forge.R-project.org", "http://cran.cnr.berkeley.edu"), type="source")
require(tlocoh)
```

The following code will estimate the local convex hull estimate range of animals of day of year. 

```{r getLoCoH}
individuals_locoh <- caribou_sf %>%
  ddply(c("ID","Year"), getLoCoH, nn = 30) 

#head(individuals_locoh)

# visualize
mapview::mapview(individuals_locoh %>% st_as_sf, zcol = "ID")
```


The default number of nearest neighbors in the `getLoCoH` function as been set to 10. However, it is  worth noting that the resulting area is sensitive to the number of nearest neighbors chosen when computing the Local Convex Hulls (LoCoH). We recommend trying different numbers of nearest neighbors to compute ranges.

Let's try here with a bigger number of nearest neighbors:
```{r getLoCoH2}
individuals_locoh2 <- caribou_daily_mean %>% st_as_sf(coords=c("x", "y"), crs = 32610) %>%
  ddply(c("ID","Year"), getLoCoH, nn = 30) %>% st_as_sf(crs = 32610)
head(individuals_locoh2)

#  visualize
mapview::mapview(individuals_locoh2, zcol = "ID")
```

We can see that the more nearest neighbors, the less sensitive the method to areas that are actively avoided. 
